```{r env, echo=FALSE}
opts_chunk$set(comment=NA, echo=FALSE)
```

`MSnbase` is an R/Bioconductor package that provides infrastructure
for plotting, data manipulation and processing of MS-based proteomics
data. The project was started by
[Laurent Gatto](http://lgatto.github.io/) in October 2010 (Mon Oct 4
23:35:23 2010, according to the git log) and has, since then,
benefited from various contributions. 

```{r}
auth <- packageDescription("MSnbase")$Author
ctrb <- sub("<lg390@cam.ac.uk> ", "",
            gsub("\t", " ", gsub("\n", " ", auth)))
cat(gsub("  ", " ", ctrb), "\n")
```

The official page is the Bioconductor landing page
([release](http://www.bioconductor.org/packages/release/bioc/html/MSnbase.html)
or
[devel](http://www.bioconductor.org/packages/devel/bioc/html/MSnbase.html)
versions). The [github page](https://github.com/lgatto/MSnbase) page
is for sharing, testing, issue tracking and forking/pulling
purposes. The
[Bioconductor svn server](https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/),
is the official repository for the working source code.


To cite the MSnbase package in publications, please use:

> Gatto L, Lilley KS. *`MSnbase` - an R/Bioconductor package for
> isobaric tagged mass spectrometry data visualization, processing and
> quantitation*. Bioinformatics. 2012 Jan
> 15;28(2):288-9. doi:[10.1093/bioinformatics/btr645](http://bioinformatics.oxfordjournals.org/content/28/2/288.long). Epub
> 2011 Nov 22.  PubMed
> [PMID:22113085](http://www.ncbi.nlm.nih.gov/pubmed/22113085).


