---
title: "MSnbase: centroiding of profile-mode MS data"
author: 
- name: Johannes Rainer
package: MSnbase
abstract: >
 This vignette describes the use of the MSnbase package for centroiding of
 profile-mode mass spectrometry data.
output:
  BiocStyle::html_document:
    toc_float: true
bibliography: MSnbase.bib	
vignette: >
  %\VignetteIndexEntry{MSnbase: centroiding of profile-mode MS data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteKeywords{Mass Spectrometry, MS, MSMS, Proteomics, Infrastructure, Bioinformatics}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'asis'}
library(BiocStyle)
BiocStyle::markdown()
```

# Introduction

Profile raw data detected by a mass spectromer consists of a distributed signal
across m/z values at each point where an ion is detected (which depends on the
resolution and settings of the instrument) [@Smith:2014di]. Profile mode data
can be processed into *centroid* data by retaining only the local maximum in
each range in which an ion is detected. This *centroiding* largely reduces the
amount of data without much loss in information content. Certain algorithms,
such as the *centWave* method in the `r Biocpkg("xcms")` package for
chromatographic peak detection in LC-MS experiments, require the data to be in
centroid mode.
 
For mass spectrometry systems from many manufacturers centroiding can already be
performed within the vendor's software during or after data acquisition. The
software from some vendors does however generate centroided data of poor
quality. `MSnbase` provides some base functionality to perform centroiding of
profile MS data, and, with the possibility of exporting the centroided MS data
to *mzML* files, the results of this centroiding can also be used as input for
other software, if needed.

# Centroiding of profile-mode MS data

In this vignette we use a subset of a profile-mode LC-MS data of pooled human
serum samples measured on a AB Sciex TripleTOF 5600+ mass spectrometer (the
employed chromatography was a hydrophilic interaction high-performance liquid
chromatography (HILIC HPLC)). The mzML file contains profile mode data for an
m/z range from 105 to 130 and a retention time from 0 to 240 seconds. For more
details on the sample see `?msdata::sciexdata`. Below we load the required
libraries and read the MS data.

```{r libraries, message = FALSE}
library(MSnbase)
library(msdata)

fl <- dir(system.file("sciex", package = "msdata"), full.names = TRUE)[2]
data_prof <- readMSData(fl, mode = "onDisk", centroided = FALSE)
```

We next extract the profile MS data for the [M+H]+ adduct of serine with the
expected m/z of 106.049871. We thus filter the `data_prof` object using an m/z
range containing the signal for the metabolite and a retention time window from
175 to 187 seconds corresponding to the time when the analyte elutes from the
LC. Subsequently we plot the profile MS data for serine.

```{r serine-msdata, fig.align = "center", fig.cap = "MS profile data for serine. Upper panel shows the base peak chromatogram (BPC), lower panel the individual signals in the retention time - m/z space. The horizontal dashed red line indicates the theoretical m/z of the [M+H]+ adduct of serine."}
## Define the mz and retention time ranges
serine_mz <- 106.049871
mzr <- c(serine_mz - 0.01, serine_mz + 0.01)
rtr <- c(175, 187)

## Filtering the object
serine <- filterMz(filterRt(data_prof, rt = rtr), mz = mzr)

plot(serine, type = "XIC")
abline(h = serine_mz, col = "red", lty = 2)
```

The lower panel in the plot above shows the individual signal intensities
measured by the mass spectrometer, the upper part the *base peak chromatogram*
(BPC), which represents the maximum signal for each discrete retention time
across the range of m/z values. The horizontal rows of points in the lower panel
indicate the resolution of the mass spectrometer. Each column of data points
(for a discrete time point) represents the signal for the ion in one
spectrum. Below we plot the signal for one of the spectra containing signal for
serine.

```{r serine-spectrum24, fig.align = "center", fig.cap = "On of the spectra for serine in profile mode."}
plot(serine[[22]])
```

The MS instrument recorded a signal along the m/z range in discrete intervals
(which depend on the resolution of the instrument). The profile-mode signal of
the serine ion at the respective retention time (the *mass peak*) consists
therefore of multiple intensities that follow approximately a gaussian
distribution.

As detailed in the introduction, centroiding aims to reduce this signal
distribution to a single representative intensity for the ion in a spectrum
(i.e. at a given retention time). The simplest approach selects the largest
intensity for each mass peak and report its intensity and m/z value. This can be
done using the `pickPeaks` method with default parameters as shown below.

```{r simple-pickPeaks, fig.align = "center", warning = FALSE, fig.cap = "Centroided data for serine."}

data_cent <- pickPeaks(data_prof)

## Plot the centroided data for serine
plot(filterMz(filterRt(data_cent, rtr), mzr), type = "XIC")
abline(h = serine_mz, col = "red", lty = 2)

```

After centroiding the data consists of a single intensity for each
mass peak. In the example above the centroids from consecutive scans do however
not have the same m/z value, but they *jump* between the discrete values defined
by the mass spectrometry's resolution. For lower intensity signals this jumping
can be even stronger. To illustrate this we plot below the centroided signal for
the [M+H]+ ion of proline.

```{r proline, fig.align = "center", warning = FALSE, fig.cap = "Centroided data for proline."}
prol_mz <- 116.070608
prol_mzr <- c(prol_mz - 0.01, prol_mz + 0.01)
prol_rtr <- c(165, 175)

plot(filterMz(filterRt(data_cent, prol_rtr), prol_mzr), type = "XIC")
abline(h = prol_mz, col = "red", lty = 2)

```

For proline the centroids jump between 3 bins of m/z values in consecutive
scans. Also the chromatographic data does not show a nice peak shape. Additional
data processing, such as data smoothing prior to centroiding and/or *refining*
the centroid's m/z can reduce these effects and improve overall data quality as
we will see in the next section.

# Improving the signal quality

While the simple centroiding using `pickPeaks` as described in the previous
section might be sufficient for many experiments and setups, MS data smoothing
and refinement of the identified centroids' m/z values can improve data quality.

## Data smoothing {#sec:smoothing}

Raw mass spectrometry data is usually smoothed in m/z dimension by applying
e.g. a Savitzky-Golay filter [@Savitzky:1964bn] which reduces the noise and
hence improves data quality. Below we use the `smooth` method to apply a
Savitzky-Golay filter with a half-window size of 4 to the data within each
spectrum (see `?smooth` for more details on the parameters).

```{r smoothSG}
data_sg <- smooth(data_prof, method = "SavitzkyGolay", halfWindowSize = 4L)

```

We next apply the simple peak picking on the smoothed data and subsequently plot
the such centroided data for serine.

```{r smoothSG-pp-serine, warning = FALSE, fig.align = "center", fig.cap = "Centroided data for serine after smoothing with a Savitzky-Golay filter."}

data_sg_cent <- pickPeaks(data_sg)

## Plot the centroided data for serine
plot(filterMz(filterRt(data_sg_cent, rtr), mzr), type = "XIC")
abline(h = serine_mz, col = "red", lty = 2)
```

Smoothing the raw data prior to peak picking improved the quality of the
centroided data of serine (and also the data for proline as can be seen below).

```{r smoothSG-pp-proline, warning = FALSE, fig.align = "center", fig.cap = "Centroided data for proline after smoothing with a Savitzky-Golay filter."}
plot(filterMz(filterRt(data_sg_cent, prol_rtr), prol_mzr), type = "XIC")
abline(h = prol_mz, col = "red", lty = 2)

```

The smoothed centroided data for proline still shows some jumping of m/z
values. Also the chromatographic data is not perfect.

In addition to smoothing the signal in m/z dimension we can smooth the signal
also in retention time dimension using the `combineSpectraMovingWindow`
function. This function aggregates signal for the same m/z value from
neighboring spectra in a moving window approach thus smoothing the
chromatographic data (by replacing the intensity in the middle spectrum by the
average signal of all intensities for the m/z in the spectra within the defined
window). To reduce the run-time of the example we apply the smoothing only to
the profile-mode data for a retention time window containing proline (in a real
data analysis this should be performed on the full data).

```{r rtsmooth, message = FALSE, warning = FALSE}
## Subset to the data for proline.
proline <- filterRt(data_prof, prol_rtr)

## Smooth the data in rt dimension
proline_c <- combineSpectraMovingWindow(proline)

## Perform the centroiding
proline_c_cent <- pickPeaks(proline_c)
```


```{r proline-rtsmooth, message = FALSE, warning = FALSE, fig.align = "center", fig.cap = "Centroided data for proline after smoothing in retention time dimension."}

plot(filterMz(filterRt(proline_c_cent, prol_rtr), prol_mzr), type = "XIC")
abline(h = prol_mz, col = "red", lty = 2)
```

Smoothing in retention time dimension improves the chromatographic peak shape of
proline.

Note however that, to combine data from multiple spectra, the
`combineSpectraMovingWindow` function has to first load the full data into
memory (i.e. it converts the `OnDiskMSnExp` object into a `MSnExp` object) and
that it returns also a `MSnExp` object. In a real use case it is thus advisable
to apply `combineSpectraMovingWindow` separately on each file of an experiment
and to export the results as an *mzML* file using the `writeMSData` method.

## Refinement of the centroid's m/z values {#sec:refine}

Thus far we applied only a simple peak picking strategy, but the `pickPeaks`
method allows also to *refine* the identified centroid's m/z value by
considering also the signal from the full, or parts of the, mass peak. Currently
two methods are implemented, *descendPeak* and *kNeighbors* that can be selected
by passing either `refineMz = "descendPeak"` or `refineMz = "kNeighbors"` to the
`pickPeaks` method. The m/z value of the reported centroid is calculated using
an intensity-weighted mean of m/z-intensity values from the mass peak. This can
improve the accuracy of the reported m/z values. The two methods differ only in
the way in which the peak area for the final calculation is defined:
`kNeighbors` simply takes the `k` m/z-intensity pairs (default `k = 2`) left and
right of the centroid and `descendPeak` walks, on both sides from the centroid,
down until the signal is below `signalPercentage`% of the centroid's intensity
(by default 33%), or until the signal increases again. All m/z intensity pairs
within this range are used for the weighted average calculation of the
centroid's m/z value.

```{r refineMz}
## Use pickPeaks with descendPeak m/z refinement
data_sg_cent_mz <- pickPeaks(data_sg, refineMz = "descendPeak")

```

Below we first extract the data for serine and then plot the smoothed and
centroided data without and with m/z refinement.

```{r refineMz-serine, message = FALSE, warning = FALSE, fig.align = "center", fig.cap = "Smoothed and centroided data for serine without (left) and with m/z refinement (right). The horizontal red dashed line indicates the theoretical m/z for the [M+H]+ ion and the vertical red dotted line the position of the largest signal for serine."}
## Extract the data for serine
serine_sg_cent <- filterMz(filterRt(data_sg_cent, rtr), mzr)
serine_sg_cent_mz <- filterMz(filterRt(data_sg_cent_mz, rtr), mzr)

## Plot the data
layout(matrix(1:4, ncol = 2))
## No m/z refinement
plot(serine_sg_cent, type = "XIC", layout = NULL)
abline(h = serine_mz, col = "red", lty = 2)
abline(v = rtime(serine_sg_cent)[22], col = "red", lty = 3)
## With m/z refinement
plot(serine_sg_cent_mz, type = "XIC", layout = NULL)
abline(h = serine_mz, col = "red", lty = 2)
abline(v = rtime(serine_sg_cent_mz)[22], col = "red", lty = 3)

```

While the centroided data with m/z refinement looks slightly noisier, the
accuracy was improved as shown by the difference between the largest signal
centroids' m/z value and the theoretical m/z value for the [M+H]+ ion of serine.

For the simple peak picking on raw data the difference is:

```{r serine-diff-mz-raw, warning = FALSE, message = FALSE}
## only centroided
mz(filterMz(filterRt(data_cent, rtr), mzr))[[22]] - serine_mz

```

Smoothing already improves the accuracy:

```{r serine-diff-mz-sg, warning = FALSE, message = FALSE}
## smoothed and centroided
mz(serine_sg_cent)[[22]] - serine_mz

```

And refining the m/z value during the centroiding can improve accuracy even
more:

```{r serine-diff-mz-sg-ref, warning = FALSE, message = FALSE}
## smoothed and centroided with m/z refinement
mz(serine_sg_cent_mz)[[22]] - serine_mz

```

Also for proline, m/z refinement improved the accuracy.

```{r refineMz-proline, message = FALSE, warning = FALSE, fig.align = "center", fig.cap = "Smoothed and centroided data for proline without (left) and with m/z refinement (right). The horizontal red dashed line indicates the theoretical m/z for the [M+H]+ ion and the vertical red dotted line the position of the maximum signal."}

proline_sg_cent <- filterMz(filterRt(data_sg_cent, prol_rtr), prol_mzr)
proline_sg_cent_mz <- filterMz(filterRt(data_sg_cent_mz, prol_rtr), prol_mzr)

layout(matrix(1:4, ncol = 2))
plot(proline_sg_cent, type = "XIC", layout = NULL)
abline(h = prol_mz, col = "red", lty = 2)
abline(v = rtime(proline_sg_cent_mz)[16], col = "red", lty = 3)

plot(proline_sg_cent_mz, type = "XIC", layout = NULL)
abline(h = prol_mz, col = "red", lty = 2)
abline(v = rtime(proline_sg_cent_mz)[16], col = "red", lty = 3)
```

The difference between the m/z of the centroid with the largest signal and the
theoretical m/z for the [M+H]+ ion of proline is shown below.

```{r proline-diff-mz, warning = FALSE, message = FALSE}
## only centroiding
mz(filterMz(filterRt(data_cent, prol_rtr), prol_mzr))[[16]] - prol_mz

## smoothed and centroided
mz(proline_sg_cent)[[16]] - prol_mz

## smoothed and centroided with m/z refinement
mz(proline_sg_cent_mz)[[16]] - prol_mz

```

Summarizing, smoothing raw profile MS data, e.g. by applying a Savitzky-Golay
filter, improves quality considerably. Additional smoothing in retention time
dimension can be advantageous too, specifically for the chromatographic peak
shape. Accuracy can be further improved for smoothed profile MS data by refining
the m/z value of the identified centroids.

# References
